shorts: An R Package for Modeling Short Sprints

### 2021-07-21 paper.tex
Added section to want about predict_XXX issues with distance correction model when distance correction <>0m

\documentclass[fleqn,10pt,lineno]{wlpeerj} % for journal submissions

@@

% Pandoc citation processing
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc} % for calculating minipage widths
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

The \textbf{shorts} package (\protect\hyperlink{ref-R-shorts}{Jovanovic 2020}), written in the R language (\protect\hyperlink{ref-R-base}{R Core Team 2020}), represents an open-source tool to help sports scientists translate raw timing data into detailed mechanical outcomes through mathematical modeling (\protect\hyperlink{ref-morinSimpleMethodComputing2019}{Jean-Benoit Morin et al. 2019}; \protect\hyperlink{ref-samozinoSimpleMethodMeasuring2016}{Samozino et al. 2016}). To best of our knowledge, scientist, researchers, and coaches have been performing short sprints modeling using the built-in solver function of Excel (Microsoft Corporation, Redmond, Washington, United States) (\protect\hyperlink{ref-morinSpreadsheetSprintAcceleration2017}{J. B. Morin 2017}; \protect\hyperlink{ref-morinSpreadsheetSprintAcceleration2019}{Jean-Benoit Morin and Samozino 2019}; \protect\hyperlink{ref-stenrothForcevelocityProfilingIce2020}{Stenroth, Vartiainen, and Karjalainen 2020}; \protect\hyperlink{ref-stenrothSpreadsheetSprintAcceleration2020}{Stenroth and Vartiainen 2020}; \protect\hyperlink{ref-samozinoSimpleMethodMeasuring2016}{Samozino et al. 2016}; \protect\hyperlink{ref-clarkNFLCombine40Yard2017}{Clark et al. 2017}; \protect\hyperlink{ref-morinSimpleMethodComputing2019}{Jean-Benoit Morin et al. 2019}), which makes the \textbf{shorts} package a major improvement in ease-of-use, speed, transparency, reproducibility, and more feature-rich model fitting.

@@

@@

@@

Short sprints profiling is usually performed by: (1) measuring split times using timing gates (i.e., positioned at various distances, e.g., 5, 10, 20, 30, 40 m), (2) getting a velocity trace using a radar gun. Estimation of MSS and TAU parameters from equation \eqref{eq:velocity-time} is performed in \textbf{shorts} package using non-linear least squares regression implemented in the \texttt{nls()} function in the \textbf{base R} (\protect\hyperlink{ref-R-base}{R Core Team 2020}) and \texttt{nlme()} function in the \textbf{nlme} package (\protect\hyperlink{ref-R-nlme}{Pinheiro, Bates, and R-core 2020}) for the mixed-effect models.

\hypertarget{estimating-short-sprint-parameters-using-split-times}{%
\subsection{Estimating short sprint parameters using split times}\label{estimating-short-sprint-parameters-using-split-times}}

Let's consider an example of an athlete with MSS equal to 9 \(ms^{-1}\), TAU equal to 1.3, and MAC equal to 6.92 \(ms^{-2}\) performing 40m sprint with timing gates positioned at each 10m split. For split times, distance is a predictor, and time is the outcome variable, thus the equation \eqref{eq:velocity-time} becomes:

@@

\(W\) in equation \eqref{eq:time-distance} represents Lambert's W function (\protect\hyperlink{ref-R-LambertW}{Goerg 2020}). MSS and TAU parameters are estimated using \texttt{model\_using\_splits()} function:

@@

\NormalTok{m1}
\CommentTok{\#\textgreater{} Estimated model parameters}
\CommentTok{\#\textgreater{} {-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}}
\CommentTok{\#\textgreater{}                 MSS                 TAU                 MAC }
\CommentTok{\#\textgreater{}                9.01                1.31                6.89 }
\CommentTok{\#\textgreater{}                PMAX     time\_correction distance\_correction }
\CommentTok{\#\textgreater{}               15.52                0.00                0.00 }
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Model fit estimators}
\CommentTok{\#\textgreater{} {-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}}
\CommentTok{\#\textgreater{}       RSE R\_squared    minErr    maxErr maxAbsErr      RMSE }
\CommentTok{\#\textgreater{}   0.00249   1.00000  {-}0.00178   0.00265   0.00265   0.00176 }
\CommentTok{\#\textgreater{}       MAE      MAPE }
\CommentTok{\#\textgreater{}   0.00157   0.04742}
\end{Highlighting}
\end{Shaded}

@@

To create a simple plot of the model, use S3 \texttt{plot()} method, which returns \textbf{ggplot2} (\protect\hyperlink{ref-R-ggplot2}{Wickham, Chang, et al. 2020}) object:

@@

Plotting the model predictions can be done once we convert data from wide to long with the help of \textbf{ggplot2} (\protect\hyperlink{ref-R-ggplot2}{Wickham, Chang, et al. 2020}), \textbf{dplyr} (\protect\hyperlink{ref-R-dplyr}{Wickham, François, et al. 2020}), \textbf{tidyr} (\protect\hyperlink{ref-R-tidyr}{Wickham 2020}), and \textbf{tidyverse} (\protect\hyperlink{ref-R-tidyverse}{Wickham 2019}) packages:

@@

\NormalTok{m2}
\CommentTok{\#\textgreater{} Estimated fixed model parameters}
\CommentTok{\#\textgreater{} {-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}}
\CommentTok{\#\textgreater{}                 MSS                 TAU                 MAC }
\CommentTok{\#\textgreater{}               8.065               0.655              12.309 }
\CommentTok{\#\textgreater{}                PMAX     time\_correction distance\_correction }
\CommentTok{\#\textgreater{}              24.818               0.000               0.000 }
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Estimated random model parameters}
\CommentTok{\#\textgreater{} {-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}}
\CommentTok{\#\textgreater{}     athlete  MSS   TAU  MAC PMAX time\_correction}
\CommentTok{\#\textgreater{} 1     James 9.69 0.847 11.4 27.7               0}
\CommentTok{\#\textgreater{} 2       Jim 7.83 0.505 15.5 30.4               0}
\CommentTok{\#\textgreater{} 3      John 7.78 0.727 10.7 20.8               0}
\CommentTok{\#\textgreater{} 4 Kimberley 8.57 0.802 10.7 22.9               0}
\CommentTok{\#\textgreater{} 5  Samantha 6.45 0.395 16.3 26.4               0}
\CommentTok{\#\textgreater{}   distance\_correction}
\CommentTok{\#\textgreater{} 1                   0}
\CommentTok{\#\textgreater{} 2                   0}
\CommentTok{\#\textgreater{} 3                   0}
\CommentTok{\#\textgreater{} 4                   0}
\CommentTok{\#\textgreater{} 5                   0}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Model fit estimators}
\CommentTok{\#\textgreater{} {-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}}
\CommentTok{\#\textgreater{}       RSE R\_squared    minErr    maxErr maxAbsErr      RMSE }
\CommentTok{\#\textgreater{}    0.0260    0.9998   {-}0.0293    0.0496    0.0496    0.0214 }
\CommentTok{\#\textgreater{}       MAE      MAPE }
\CommentTok{\#\textgreater{}    0.0172    0.9019}
\end{Highlighting}
\end{Shaded}

Additional information about mixed-effects model performed using the \texttt{nlme} package (\protect\hyperlink{ref-R-nlme}{Pinheiro, Bates, and R-core 2020}) can be obtained using \texttt{summary()} function. S3 method \texttt{coef()} when applied on mixed-model result will return both the fixed and random effects. To create a simple plot of the model, use S3 \texttt{plot()} method.

@@

The following three models estimate MSS and TAU from the three datasets. Tables are created using the \texttt{kable()} function from the \textbf{knitr} package (\protect\hyperlink{ref-knitr2014}{Xie 2014}, \protect\hyperlink{ref-knitr2015}{2015}, \protect\hyperlink{ref-R-knitr}{2020}) and \texttt{kable\_styling()} function from the \textbf{kableExtra} package (\protect\hyperlink{ref-R-kableExtra}{Zhu 2021}).

@@

\hypertarget{problems-at-the-start-when-using-split-times}{%
\subsection{Problems at the start when using split times}\label{problems-at-the-start-when-using-split-times}}

@@

John represents the \emph{theoretical model}, in which we assume that the initial force production and the timing initiation are perfectly synchronized. Jack, on the other hand, represents a \emph{practical model}, and decides to move slightly behind the initial timing gate (i.e.~for 0.5 m) and use body rocking to initiate the sprint start. In other words, Jack is using a \emph{flying start}, a common scenario when testing field sports athletes.

@@

Here is a graphical representation of the sprint splits (please refer to the Supplemental files for the R code):

@@

To explore this further, we have run a simple simulation with 5, 10, 20, 30, 40, and 50 m splits, with MSS and MAC varying from 6 to 9 (\(ms^{-1}\) and \(ms^{-2}\) respectively), and flying start distance varying from 0 to 1 m (please refer to the Supplemental files for the R code). With a simulation dataset, we can check the model estimates and predictions, given the flying start distance.

The following figure demonstrates the effect of flying start distance on estimated MSS:

@@

As can be seen from the figure, MSS is underestimated as flying start distance increases. The following image demonstrates the effect of flying start distance on estimated MAC:

@@

MAC (and also TAU) are highly affected by the flying start distance, and from the figure we can notice that MAC is overestimated as flying start distance increases.

And finally, the following image demonstrates the effect of flying start distance on estimated PMAX:

@@

Estimated PMAX is also overestimated as flying start distance increases.

Model residuals are also affected by flying start distance. The shape of residuals distribution depends on number and splits utilized (e.g., 10, 20, 30, 40 m versus 5, 15, 30 m), but here we can see the effect of the flying start distance on the model residuals per split distance utilized in our simulation:

\begin{center}\includegraphics[width=0.9\linewidth]{paper_files/figure-latex/unnamed-chunk-32-1} \end{center}

If we merge individual facets (i.e., combinations of MSS and MAC), we can get simpler figure conveying issues with residuals when there is a flying start:

\begin{center}\includegraphics[width=0.9\linewidth]{paper_files/figure-latex/unnamed-chunk-33-1} \end{center}

@@

In Jack's case, both +0.3 s fixed time correction and time correction estimation yield parameters closer to John's (i.e.~true parameters).

Another model definition, which is a novel approach implemented in the \textbf{shorts} package, is to utilize \emph{distance correction}, besides time correction. Thus, equation \eqref{eq:time-distance} becomes:

\begin{equation}
  t(d) = TAU \times W(-e^{\frac{-d + distance \; correction}{MSS \times TAU}} - 1) + \frac{d + distance \; correction}{MSS} + TAU - time \; correction \label{eq:distance-correction}
\end{equation}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{jack\_profile\_distance\_correction }\OtherTok{\textless{}{-}} \FunctionTok{model\_using\_splits\_with\_corrections}\NormalTok{(}
  \AttributeTok{distance =}\NormalTok{ split\_times}\SpecialCharTok{$}\NormalTok{distance,}
  \AttributeTok{time =}\NormalTok{ split\_times}\SpecialCharTok{$}\NormalTok{jack\_time}
\NormalTok{)}

\NormalTok{jack\_parameters }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(}
  \FunctionTok{coef}\NormalTok{(john\_profile),}
  \FunctionTok{coef}\NormalTok{(jack\_profile),}
  \FunctionTok{coef}\NormalTok{(jack\_profile\_fixed\_time\_short),}
  \FunctionTok{coef}\NormalTok{(jack\_profile\_fixed\_time\_long),}
  \FunctionTok{coef}\NormalTok{(jack\_profile\_time\_estimated),}
  \FunctionTok{coef}\NormalTok{(jack\_profile\_distance\_correction)}
\NormalTok{)}

\FunctionTok{rownames}\NormalTok{(jack\_parameters) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}
  \StringTok{"John"}\NormalTok{,}
  \StringTok{"Jack {-} No corrections"}\NormalTok{,}
  \StringTok{"Jack {-} Fixed time correction (+0.3s)"}\NormalTok{,}
  \StringTok{"Jack {-} Fixed time correction (+0.5s)"}\NormalTok{,}
  \StringTok{"Jack {-} Estimated time correction"}\NormalTok{,}
  \StringTok{"Jack {-} Estimated distance correction"}
\NormalTok{)}

@@

\begin{table}[!h]
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}{lrrrrrr}
\toprule
  & MSS & TAU & MAC & PMAX & time\_correction & distance\_correction\\
\midrule
John & 9.00 & 1.30 & 6.92 & 15.6 & 0.00 & 0.0\\
Jack - No corrections & 8.49 & 0.70 & 12.06 & 25.6 & 0.00 & 0.0\\
Jack - Fixed time correction (+0.3s) & 9.00 & 1.25 & 7.19 & 16.2 & 0.30 & 0.0\\
Jack - Fixed time correction (+0.5s) & 9.62 & 1.77 & 5.43 & 13.1 & 0.50 & 0.0\\
Jack - Estimated time correction & 8.96 & 1.22 & 7.37 & 16.5 & 0.28 & 0.0\\
\addlinespace
Jack - Estimated distance correction & 9.00 & 1.30 & 6.92 & 15.6 & 0.40 & 0.5\\
\bottomrule
\end{tabular}}
\end{table}

As can be seen from the results, adding distance correction results in correctly estimating Jack's sprint parameters. There are a few issues with this model definition. Besides being novel and still not validated with actual data, distance correction model has four parameters to estimate, which implies that at least five sprint splits are needed. This imposes practical limitations, since acquiring six timing gate (one for the start and five for splits) might be practically troublesome. One strategy that is sometimes implemented is adding zeros to the sample (i.e., \(t=0\) and \(d=0\)), which increase the number of observations. Unfortunately, this strategy should not be implemented, as explained later in the \protect\hyperlink{should-we-add-zero-to-the-sample}{Should we add zero to the sample?} section of this paper.

We will get back to these issues later, but we can examine how these models (no correction, fixed time correction, estimated time correction, and estimated time and distance correction) perform using simulated data with varying flying start distance (please refer to the Supplemental files for the R code).

>>>>

\documentclass[fleqn,10pt]{wlpeerj} % for preprint submissions

@@

% Pandoc citation processing
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc} % for calculating minipage widths
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

The \textbf{shorts} package (\protect\hyperlink{ref-R-shorts}{Jovanović 2021}), written in the R language (\protect\hyperlink{ref-R-base}{R Core Team 2020}), represents an open-source tool to help sports scientists translate raw timing data into detailed mechanical outcomes through mathematical modeling (\protect\hyperlink{ref-morinSimpleMethodComputing2019}{Jean-Benoit Morin et al. 2019}; \protect\hyperlink{ref-samozinoSimpleMethodMeasuring2016}{Samozino et al. 2016}). To best of our knowledge, scientist, researchers, and coaches have been performing short sprints modeling using the built-in solver function of Excel (Microsoft Corporation, Redmond, Washington, United States) (\protect\hyperlink{ref-morinSpreadsheetSprintAcceleration2017}{J. B. Morin 2017}; \protect\hyperlink{ref-morinSpreadsheetSprintAcceleration2019}{Jean-Benoit Morin and Samozino 2019}; \protect\hyperlink{ref-stenrothForcevelocityProfilingIce2020}{Stenroth, Vartiainen, and Karjalainen 2020}; \protect\hyperlink{ref-stenrothSpreadsheetSprintAcceleration2020}{Stenroth and Vartiainen 2020}; \protect\hyperlink{ref-samozinoSimpleMethodMeasuring2016}{Samozino et al. 2016}; \protect\hyperlink{ref-clarkNFLCombine40Yard2017}{Clark et al. 2017}; \protect\hyperlink{ref-morinSimpleMethodComputing2019}{Jean-Benoit Morin et al. 2019}), which makes the \textbf{shorts} package a major improvement in ease-of-use, speed, transparency, reproducibility, and more feature-rich model fitting.

@@

\small

@@

\normalsize

@@

Short sprints profiling is usually performed by: (1) measuring split times using timing gates (i.e., positioned at various distances, e.g., 5, 10, 20, 30, 40 m), (2) getting a velocity trace using a radar gun. Estimation of MSS and TAU parameters from equation \eqref{eq:velocity-time} is performed in \textbf{shorts} package using non-linear least squares regression implemented in the \texttt{nls()} function (\protect\hyperlink{ref-batesNonlinearModels1992}{D. M. Bates and Chambers 1992}; \protect\hyperlink{ref-batesNonlinearRegressionAnalysis2007}{Douglas M. Bates and Watts 2007}) in the \textbf{base R} (\protect\hyperlink{ref-R-base}{R Core Team 2020}) and \texttt{nlme()} function in the \textbf{nlme} package (\protect\hyperlink{ref-R-nlme}{Pinheiro, Bates, and R-core 2021}) for the mixed-effect models.

\hypertarget{estimating-short-sprint-parameters-using-timing-gates-split-times}{%
\subsection{Estimating short sprint parameters using timing gates split times}\label{estimating-short-sprint-parameters-using-timing-gates-split-times}}

Let's consider an example of an athlete with MSS equal to 9 \(ms^{-1}\), TAU equal to 1.3, and MAC equal to 6.92 \(ms^{-2}\) performing 40m sprint with timing gates positioned at each 10m split. For split times, distance is a predictor, and time is the outcome variable, thus the equation \eqref{eq:distance-time} becomes:

@@

\(W\) in equation \eqref{eq:time-distance} represents Lambert's W function (\protect\hyperlink{ref-R-LambertW}{Goerg 2020}). Researchers often incorrectly use the equation \eqref{eq:distance-time} (\protect\hyperlink{ref-morinSpreadsheetSprintAcceleration2017}{J. B. Morin 2017}; \protect\hyperlink{ref-morinSpreadsheetSprintAcceleration2019}{Jean-Benoit Morin and Samozino 2019}; \protect\hyperlink{ref-stenrothSpreadsheetSprintAcceleration2020}{Stenroth and Vartiainen 2020}), in which the time is the predictor and distance is the outcome variable, instead of statistically correct equation \eqref{eq:time-distance} (\protect\hyperlink{ref-motulskyIntuitiveBiostatisticsNonmathematical2018}{Motulsky 2018, 341}).

MSS and TAU parameters are estimated using \texttt{model\_using\_splits()} function:

\small

@@

\NormalTok{m1}
\CommentTok{\#\textgreater{} Estimated model parameters}
\CommentTok{\#\textgreater{} {-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}}
\CommentTok{\#\textgreater{}                 MSS                 TAU                 MAC }
\CommentTok{\#\textgreater{}                9.01                1.31                6.89 }
\CommentTok{\#\textgreater{}                PMAX     time\_correction distance\_correction }
\CommentTok{\#\textgreater{}               15.52                0.00                0.00 }
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Model fit estimators}
\CommentTok{\#\textgreater{} {-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}}
\CommentTok{\#\textgreater{}       RSE R\_squared    minErr    maxErr maxAbsErr      RMSE }
\CommentTok{\#\textgreater{}   0.00249   1.00000  {-}0.00265   0.00178   0.00265   0.00176 }
\CommentTok{\#\textgreater{}       MAE      MAPE }
\CommentTok{\#\textgreater{}   0.00157   0.04742}
\end{Highlighting}
\end{Shaded}

\normalsize

@@

To create a simple plot of the model, use S3 \texttt{plot()} method, which returns \textbf{ggplot2} (\protect\hyperlink{ref-R-ggplot2}{Wickham, Chang, et al. 2021}) object:

\small

@@

\normalsize

Plotting the model predictions can be done once we convert data from wide to long with the help of \textbf{ggplot2} (\protect\hyperlink{ref-R-ggplot2}{Wickham, Chang, et al. 2021}), \textbf{dplyr} (\protect\hyperlink{ref-R-dplyr}{Wickham, François, et al. 2021}), \textbf{tidyr} (\protect\hyperlink{ref-R-tidyr}{Wickham 2021a}), and \textbf{tidyverse} (\protect\hyperlink{ref-R-tidyverse}{Wickham 2021b}) packages:

\small

@@

\NormalTok{m2}
\CommentTok{\#\textgreater{} Estimated fixed model parameters}
\CommentTok{\#\textgreater{} {-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}}
\CommentTok{\#\textgreater{}                 MSS                 TAU                 MAC }
\CommentTok{\#\textgreater{}               8.065               0.655              12.309 }
\CommentTok{\#\textgreater{}                PMAX     time\_correction distance\_correction }
\CommentTok{\#\textgreater{}              24.818               0.000               0.000 }
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Estimated random model parameters}
\CommentTok{\#\textgreater{} {-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}}
\CommentTok{\#\textgreater{}     athlete  MSS   TAU  MAC PMAX time\_correction}
\CommentTok{\#\textgreater{} 1     James 9.69 0.847 11.4 27.7               0}
\CommentTok{\#\textgreater{} 2       Jim 7.83 0.505 15.5 30.4               0}
\CommentTok{\#\textgreater{} 3      John 7.78 0.727 10.7 20.8               0}
\CommentTok{\#\textgreater{} 4 Kimberley 8.57 0.802 10.7 22.9               0}
\CommentTok{\#\textgreater{} 5  Samantha 6.45 0.395 16.3 26.4               0}
\CommentTok{\#\textgreater{}   distance\_correction}
\CommentTok{\#\textgreater{} 1                   0}
\CommentTok{\#\textgreater{} 2                   0}
\CommentTok{\#\textgreater{} 3                   0}
\CommentTok{\#\textgreater{} 4                   0}
\CommentTok{\#\textgreater{} 5                   0}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Model fit estimators}
\CommentTok{\#\textgreater{} {-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}{-}}
\CommentTok{\#\textgreater{}       RSE R\_squared    minErr    maxErr maxAbsErr      RMSE }
\CommentTok{\#\textgreater{}    0.0260    0.9998   {-}0.0496    0.0293    0.0496    0.0214 }
\CommentTok{\#\textgreater{}       MAE      MAPE }
\CommentTok{\#\textgreater{}    0.0172    0.9019}
\end{Highlighting}
\end{Shaded}

\normalsize

Additional information about mixed-effects model performed using the \texttt{nlme} package (\protect\hyperlink{ref-R-nlme}{Pinheiro, Bates, and R-core 2021}) can be obtained using \texttt{summary()} function. S3 method \texttt{coef()} when applied on mixed-model result will return both the fixed and random effects.

To create a simple plot of the model, use S3 \texttt{plot()} method.

\small

@@

\normalsize

The following three models estimate MSS and TAU from the three datasets. Tables are created using the \texttt{kable()} function from the \textbf{knitr} package (\protect\hyperlink{ref-knitr2014}{Xie 2014}, \protect\hyperlink{ref-knitr2015}{2015}, \protect\hyperlink{ref-R-knitr}{2021b}) and \texttt{kable\_styling()} function from the \textbf{kableExtra} package (\protect\hyperlink{ref-R-kableExtra}{Zhu 2021}).

\small

@@

\normalsize

\hypertarget{problems-at-the-start-when-using-timing-gates}{%
\subsection{Problems at the start when using timing gates}\label{problems-at-the-start-when-using-timing-gates}}

@@

John represents the \emph{theoretical model}, in which we assume that the initial force production and the timing initiation are perfectly synchronized. Jack, on the other hand, represents a \emph{practical model}, and decides to move slightly behind the initial timing gate (i.e.~for 0.5 m) and use body rocking to initiate the sprint start. In other words, Jack is using a \emph{flying start}, a common scenario when testing field sports athletes. Flying start distance is often recommended to avoid premature triggering of the timing system by lifted knees or swinging
arms (\protect\hyperlink{ref-altmannAccuracySingleBeam2018}{Altmann et al. 2018}, \protect\hyperlink{ref-altmannDifferentStartingDistances2015}{2015}, \protect\hyperlink{ref-altmannValiditySingleBeamTiming2017}{2017}; \protect\hyperlink{ref-haugenPowerForceVelocityProfilingSprinting2020}{Thomas A. Haugen, Breitschädel, and Samozino 2020}; \protect\hyperlink{ref-haugenSprintRunningPerformance2016}{T. Haugen and Buchheit 2016}).

\small

@@

\normalsize

Here is a graphical representation of the sprint splits (please refer to the \protect\hyperlink{supplemental-material}{Supplemental material} for the R code):

\small

@@

To explore this further, we have run a simple simulation by increasing Jack's flying start distance from 0 to 1 m and depicting the estimated MSS, TAU, MAC, and PMAX parameters (please refer to the \protect\hyperlink{supplemental-material}{Supplemental material} for the R code).

\small

@@

\normalsize

As can be seen from the figure, MSS is underestimated as flying start distance increases. MAC (and also TAU) are highly affected by the flying start distance, and from the figure we can notice that MAC is overestimated as flying start distance increases. Estimated PMAX is also overestimated as flying start distance increases.

Model residuals are also affected by the flying start distance. The shape of residuals distribution depends on number and splits utilized (e.g., 10, 20, 30, 40 m versus 5, 15, 30 m), but here we can see the effect of the flying start distance on the model residuals per split distance utilized in our simple simulation:

\small

@@

\normalsize

Another way to visualize the effect of the flying start distance on split distance residuals can be found on the next figure:

\small

@@

\normalsize

@@

\normalsize

In Jack's case, both +0.3 s fixed time correction and time correction estimation yield parameters closer to John's (i.e.~true parameters). We have used these two model definitions in retrospective pilot study (\protect\hyperlink{ref-vescoviSprintMechanicalCharacteristics2021}{Vescovi and Jovanovi'c 2021}), demonstrating statistically significant differences in estimated FVP parameters.

Instead of using time correction as simple intercept in the equation \eqref{eq:time-correction}, we can estimate the \emph{distance correction} (i.e., the flying start distance) using the following equation:

\begin{equation}
  \begin{split}
   t(d) &= (TAU \times W(-e^{\frac{-d + distance \; correction}{MSS \times TAU}} - 1) + \frac{d + distance \; correction}{MSS} + TAU) \\ 
   &\quad-(TAU \times W(-e^{\frac{distance \; correction}{MSS \times TAU}} - 1) + \frac{distance \; correction}{MSS} + TAU) 
   \end{split}
   \label{eq:distance-correction}
\end{equation}

Equation \eqref{eq:distance-correction} was used to generate the data for John and Jack (please refer to the provided R code), as well as for the simple simulation, and can be used as model definition to control for the flying start distance.

To estimate distance correction parameter, we use \texttt{model\_using\_splits\_with\_distance\_correction()} function. Here is how we can estimate Jack parameters using distance correction:

\small

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{jack\_profile\_distance\_correction }\OtherTok{\textless{}{-}} \FunctionTok{model\_using\_splits\_with\_distance\_correction}\NormalTok{(}
  \AttributeTok{distance =}\NormalTok{ split\_times}\SpecialCharTok{$}\NormalTok{distance,}
  \AttributeTok{time =}\NormalTok{ split\_times}\SpecialCharTok{$}\NormalTok{jack\_time}
\NormalTok{)}

\NormalTok{jack\_parameters }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(}
\NormalTok{  jack\_parameters,}
  \FunctionTok{coef}\NormalTok{(jack\_profile\_distance\_correction)}
\NormalTok{)}

\FunctionTok{rownames}\NormalTok{(jack\_parameters) [}\DecValTok{6}\NormalTok{] }\OtherTok{\textless{}{-}} \StringTok{"Jack {-} Estimated distance correction"}

\begin{table}[!h]
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}{lrrrrrr}
\toprule
  & MSS & TAU & MAC & PMAX & time\_correction & distance\_correction\\
\midrule
John & 9.00 & 1.30 & 6.92 & 15.6 & 0.00 & 0.0\\
Jack - No corrections & 8.49 & 0.70 & 12.06 & 25.6 & 0.00 & 0.0\\
Jack - Fixed time correction (+0.3s) & 9.00 & 1.25 & 7.19 & 16.2 & 0.30 & 0.0\\
Jack - Fixed time correction (+0.5s) & 9.62 & 1.77 & 5.43 & 13.1 & 0.50 & 0.0\\
Jack - Estimated time correction & 8.96 & 1.22 & 7.37 & 16.5 & 0.28 & 0.0\\
\addlinespace
Jack - Estimated distance correction & 9.00 & 1.30 & 6.92 & 15.6 & 0.00 & 0.5\\
\bottomrule
\end{tabular}}
\end{table}

\normalsize

Another model definition, which also represents a novel approach implemented in the \textbf{shorts} package, is to utilize both time and distance corrections. Thus, equation \eqref{eq:time-correction} becomes:

\begin{equation}
  t(d) = TAU \times W(-e^{\frac{-d + distance \; correction}{MSS \times TAU}} - 1) + \frac{d + distance \; correction}{MSS} + TAU - time \; correction \label{eq:time-distance-correction}
\end{equation}

@@

\small

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{jack\_profile\_time\_distance\_correction }\OtherTok{\textless{}{-}} \FunctionTok{model\_using\_splits\_with\_corrections}\NormalTok{(}
  \AttributeTok{distance =}\NormalTok{ split\_times}\SpecialCharTok{$}\NormalTok{distance,}
  \AttributeTok{time =}\NormalTok{ split\_times}\SpecialCharTok{$}\NormalTok{jack\_time}
\NormalTok{)}

\NormalTok{jack\_parameters }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(}
\NormalTok{  jack\_parameters,}
  \FunctionTok{coef}\NormalTok{(jack\_profile\_time\_distance\_correction)}
\NormalTok{)}

\FunctionTok{rownames}\NormalTok{(jack\_parameters) [}\DecValTok{7}\NormalTok{] }\OtherTok{\textless{}{-}} 
  \StringTok{"Jack {-} Estimated time \& distance corrections"}

@@

\begin{table}[!h]
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}{lrrrrrr}
\toprule
  & MSS & TAU & MAC & PMAX & time\_correction & distance\_correction\\
\midrule
John & 9.00 & 1.30 & 6.92 & 15.6 & 0.00 & 0.0\\
Jack - No corrections & 8.49 & 0.70 & 12.06 & 25.6 & 0.00 & 0.0\\
Jack - Fixed time correction (+0.3s) & 9.00 & 1.25 & 7.19 & 16.2 & 0.30 & 0.0\\
Jack - Fixed time correction (+0.5s) & 9.62 & 1.77 & 5.43 & 13.1 & 0.50 & 0.0\\
Jack - Estimated time correction & 8.96 & 1.22 & 7.37 & 16.5 & 0.28 & 0.0\\
\addlinespace
Jack - Estimated distance correction & 9.00 & 1.30 & 6.92 & 15.6 & 0.00 & 0.5\\
Jack - Estimated time \& distance corrections & 9.00 & 1.30 & 6.92 & 15.6 & 0.40 & 0.5\\
\bottomrule
\end{tabular}}
\end{table}

\normalsize

As can be seen from the results, adding distance correction results in correctly estimating Jack's sprint parameters. There are a few issues with this model definition. Besides being novel and still not validated with actual data, distance correction model has four parameters to estimate, which implies that at least four sprint splits are needed. This imposes practical limitations, since acquiring five timing gates (one for the start and four for splits) might be practically troublesome. One strategy that is sometimes implemented is adding zeros to the sample (i.e., \(t=0\) and \(d=0\)), which increase the number of observations. Unfortunately, this strategy should not be implemented since it will bias the estimated parameters:

\small
